Generating Interactive WebSocket Applications in TypeScript

### 2020-03-09 PLACES_2020/encoding.tex
WIP Proofreading Sec. 3

Developers can implement their application using APIs generated from the EFSM
to guarantee correctness by construction.
Our approach integrates the EFSM into the development workflow by encoding
session types as TypeScript types.
Communication over the WebSocket protocol introduces additional constriants:
communication is always initiated in the front-end and driven by user interactions,
whilst back-end roles can only accept connections.
This motivates our design of encoding the session types differently for server (\cref{section:server}) and client (\cref{section:browser}) targets.

\fz{I think there is an important thing to note here, ``Server'' side is
  basically the side which accepts the connection and ``Client'' side is the
side which makes the connection. But this is orthogonal to whether it runs on
Node or on browser? (Correct me if I'm wrong)}

\am{Not for the case of WebSockets, because only WebSocket clients on browser can initiate connections and only WebSocket Server on Node can accept; will make this distinction clear.}

@@

We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
this section.
For server-side targets, we encode EFSM states into TypeScript types and
consider receive and send states separately.
Each TypeScript encoding is assigned to its state identifier from the EFSM as
its type alias, thus providing syntactic sugar when referring to the successor
state in the TypeScript encoding of the current state.
For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
encoding as $\llbracket S \rrbracket$. We outline the encoding below using examples from the
\textit{Noughts and Crosses} game (\cref{lst:svr}).

\paragraph{Branching State}
We consider a receive state as a unary branching state for conciseness.
A branching state is encoded as an \textit{object literal}
\cite{TypeScriptSpec}, with each branch $i \in I$, $I$ denoting set of all
branches, corresponding to a member field.
A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an
\textit{member field} named $\texttt{label}_i$ of function type
$(payload:\texttt{T}_i) \to \llbracket S_i \rrbracket$.
The developer implements a branching operation by passing callbacks for each
branch, parameterised by the expected message payload type for that branch.

\paragraph{Selection State}
We consider a send state as a unary selection state for conciseness.
A selection state is encoded as a \textit{union type}
\cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in
I$, $I$ denoting set of all choices, sending a message labelled
$\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state
$S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$,
  $\llbracket S_i \rrbracket$]}.
The developer implements a selection operation by passing the selected label
and payload to send in the message.
We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels
in the protocol.

@@

% REVIEW:
% Also, when describing Listing 2, it would be helpful to say a bit more explicitly that the client is supposed to implement S13, and the code in S13 returns an object of type S15. Then you can point to Listing 4 if they are coordinated.
In the case of \cref{lst:svr}, the developer is expected to implement
\texttt{S13} which handles the \texttt{Pos} message sent by P1,
and the code in \texttt{S13} returns a value of type \texttt{S15}, which corresponds to a selection of messages to send to P2. \cref{lst:svrprotocol} illustrates how the developer may implement these types.

We make a key design decision \textit{not} to expose communication channels in
the TypeScript session type encoding to provide \textit{static} linearity
guarantees (\cref{section:serverlinear}).
Our encoding sufficiently exposes seams for the developer to inject their
business logic, whilst the generated session API
(\cref{section:serversessionapi}) handles the sending and receiving of
messages.

@@

The generated code for our session runtime performs communication in a protocol-conformant manner, but
does not expose these IO actions to the developer by delegating the
aforementioned responsibilities to an inner class.
% REVIEW: it is not clear what is
% generated on the server-side and what actually is the session runtime
The runtime executes the EFSM by keeping track of 
the current state (similar to the generated code in \cite{javatypestate}) and only permitting the allowed IO actions at the current state.
The runtime listens to message (receive) events on the communication channel,
invokes the corresponding callback to obtain the value to send next, and
performs the send.
The developer instantiates the session by constructing an instance of the
session runtime class, providing the WebSocket endpoint URL (to open the
connection) and initial state (to execute the EFSM).

\subsubsection{Linear Channel Usage}
\label{section:serverlinear}
Channel linearity is guaranteed based on the correctness of our library design
and session runtime implementation, which means a faulty implementation could
violate this, but is up to the library author to verify once rather than the
end user. Our library design prevents the two conditions detailed below:

\paragraph{Repeated Usage}
We do not expose channels to the developer, which makes \textit{reuse}
impossible.
For example, to send a message, the generated API only requires the payload
that needs to be sent, and the session runtime performs the send internally,
guaranteeing this action is done \textit{exactly once} by construction.

\paragraph{Unused Channels}
The initial state must be supplied to the session runtime
constructor in order to instantiate the session;
this initial state is defined
in terms of the successor states, which in turn has references to its
successors and so forth.
This encoding approach covers the terminal state (if it exists), and the
session runtime guarantees this terminal state, if it exists, will be reached
by construction.

\subsection{The React Framework}
Our browser-side session type encodings for browser-side targets build upon the React framework, developed by Facebook \cite{React} for the \textit{Model-View-Controller} (MVC) architecture. React is widely used in industry to create scalable single-page TypeScript
applications, and we intend for our proposed workflow to be beneficial in an
industrial context.
We introduce the key features of the framework.

\paragraph{Components}
A component is a reusable UI element which
contains its own markup and logic.
Components implement a \texttt{render()} function which returns a React
element, the smallest building blocks of a React application, analogous to the
view function in the MVU architecture.
Components can keep \textit{state} and the \texttt{render()} function is
invoked upon a change of state.
A simple counter can be implemented as a component,
with \texttt{count} stored as state, a button which increments \texttt{count}
when clicked and a \texttt{div} that renders the current
\texttt{count}.
Components can also render other components, which gives rise
to a parent/child relationship between components.
Parents can pass data to children as \textit{props} (short for properties).
Going back to the aforementioned example, the counter component could
render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
its \texttt{render()} function, propagating the \texttt{count} from its state
to the child.
This enables reusability, and for our use case, gives control to the parent
on what data to pass to its children (e.g. pass the payload of a received
message to a child to render).

>>>>

Developers can implement their application using APIs generated from the EFSM
to guarantee correctness by construction.
Our approach integrates the EFSM into the development workflow by encoding
session types as TypeScript types.
Communication over the WebSocket protocol introduces additional constraints:
communication is always initiated in the front-end and driven by user interactions,
whilst back-end roles can only accept connections.
This motivates our design of encoding the session types differently for server
(\cref{section:server}) and client (\cref{section:browser}) targets.

@@

We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
this section.
For server-side targets, we encode EFSM states into TypeScript types and
consider branching (receiving) and selection (sending) states separately.
We assign TypeScript encodings of states to their state identifiers in the
EFSM, providing syntactic sugar when referring to the successor state when
encoding the current state.
For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
encoding as $\llbracket S \rrbracket$.
We outline the encoding below using examples from the
\textit{Noughts and Crosses} game (\cref{lst:svr}).

\paragraph{Branching State}
We consider a receiving state as a unary branching state for conciseness.
A branching state is encoded as an \textit{object literal}
\cite{TypeScriptSpec} (a record type), with each branch $i \in I$ ($I$ denoting set of all
branches), corresponding to a member field.
A branch expecting to receive a message labelled $\texttt{label}_i$ carrying
payload of type $\texttt{T}_i$ with successor state $S_i$ is encoded as an
\textit{member field} named $\texttt{label}_i$ of function type
$(payload:\texttt{T}_i) \to \llbracket S_i \rrbracket$.
The developer implements a branching operation by passing callbacks for each
branch, parameterised by the expected message payload type for that branch.

\paragraph{Selection State}
We consider a sending state as a unary selection state for conciseness.
A selection state is encoded as a \textit{union type}
\cite{TypeScriptSpec} of internal choice encodings: each internal choice $i \in
I$, $I$ denoting set of all choices, sending a message labelled
$\texttt{label}_i$ carrying payload of type $\texttt{T}_i$ with successor state
$S_i$ is encoded as a \textit{tuple type} of \texttt{[Labels.label$_i$, T$_i$,
  $\llbracket S_i \rrbracket$]}.
The developer implements a selection operation by passing the selected label
and payload to send in the message.
We generate a \textit{string enum} (named \texttt{Labels}) wrapping the labels
in the protocol.

@@

% REVIEW:
% Also, when describing Listing 2, it would be helpful to say a bit more explicitly that the client is supposed to implement S13, and the code in S13 returns an object of type S15. Then you can point to Listing 4 if they are coordinated.
In the case of \cref{lst:svr}, the developer is expected to implement
\texttt{S13} which handles the \texttt{Pos} message sent by \texttt{P1},
and the code in \texttt{S13} returns a value of type \texttt{S15}, which
corresponds to a selection of messages to send to \texttt{P2}. \Cref{lst:svrprotocol}
illustrates how the developer may implement these types.

We make a key design decision \textit{not} to expose communication channels in
the TypeScript session type encodings to provide \textit{static} linearity
guarantees (\cref{section:serverlinear}).
Our encoding sufficiently exposes seams for the developer to inject their
program logic, whilst the generated session API
(\cref{section:serversessionapi}) handles the sending and receiving of
messages.

@@

The generated code for our session runtime performs communication in a protocol-conformant manner, but
does not expose these IO actions to the developer by delegating the
aforementioned responsibilities to an inner class.
% REVIEW: it is not clear what is
% generated on the server-side and what actually is the session runtime
The runtime executes the EFSM by keeping track of
the current state (similar to the generated code in \cite{javatypestate})
and only permitting the specified IO actions at the current state.
The runtime listens to message (receiving) events on the communication channel,
invokes the corresponding callback to obtain the value to send next, and
performs the sending.
The developer instantiates a session by constructing an instance of the
session runtime class, providing the WebSocket endpoint URL (to open the
connection) and the initial state (to execute the EFSM).

\subsubsection{Linear Channel Usage}
\label{section:serverlinear}
\fz{This sentence is confusing and I don't know how to fix.}
Channel linearity is guaranteed based on the correctness of our library design
and session runtime implementation, which means a faulty implementation could
violate this, but is up to the library author to verify once rather than the
end user. Our library design prevents the two conditions detailed below:

\paragraph{Repeated Usage}
We do not expose channels to the developer, which makes \textit{reusing
  channels} impossible.
For example, to send a message, the generated API only requires the payload
that needs to be sent, and the session runtime performs the send internally,
guaranteeing this action is done \textit{exactly once} by construction.

\paragraph{Unused Channels}
The initial state must be supplied to the session runtime
constructor in order to instantiate a session;
this initial state is defined
in terms of the successor states, which in turn has references to its
successors and so forth.
\fz{It's not the approach, rather than the developer implementation?}
This encoding approach covers the terminal state (if it exists), and the
session runtime guarantees this terminal state, if it exists, will be reached
by construction.

\subsection{The React Framework}
Our browser-side session type encodings for browser-side targets build upon the
\emph{React.js} framework, developed by Facebook \cite{React} for the
\textit{Model-View-Controller} (MVC) architecture.
React is widely used in industry to create scalable single-page TypeScript
applications, and we intend for our proposed workflow to be beneficial in an
industrial context.
We introduce the key features of the framework.

\paragraph{Components}
A component is a reusable UI element which
contains its own markup and logic.
Components implement a \texttt{render()} function which returns a React
element, the smallest building blocks of a React application, analogous to the
view function in the MVU architecture.
Components can keep \textit{state}s and the \texttt{render()} function is
invoked upon a change of state.

For example, a simple counter can be implemented as a component,
with its \texttt{count} stored as state.
When rendered, it displays a button which increments \texttt{count}
when clicked and a \texttt{div} that renders the current
\texttt{count}.
If the button is clicked, the \texttt{count} is incremented, which triggers a
re-rendering (since the state has changed), and the updated \texttt{count} is
displayed.

Components can also render other components, which gives rise
to a parent/child relationship between components.
Parents can pass data to children as \textit{props} (short for properties).
Going back to the aforementioned example, the counter component could
render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
its \texttt{render()} function, propagating the \texttt{count} from its state
to the child.
This enables reusability, and for our use case, gives control to the parent
on what data to pass to its children (e.g. pass the payload of a received
message to a child to render).
